
---
title: 'Problem Set 05: Reshaping and Exploring Data'
author: "Your Name Here!"
output:
  html_document:
    css: "https://wjhopper.github.io/sds201_labs/assets/css/lab.css"
    theme: lumen
    toc: yes
    toc_float:
      collapsed: false
    toc_depth: 2
    df_print: paged
    self_contained: true
editor_options: 
  chunk_output_type: console
---

In this problem set, we will practice some of the key data reshaping tasks using the tools from the `tidyr` package, and practice the "getting to know you" phase of exploring a new data set with `skim` and `ggpairs`.

### Libraries
```{r warning = F, message = F}
library(ggplot2)
library(dplyr)
library(tidyr)
library(GGally)
library(skimr)
```

### The data
You will be working with the built-in `iris` data set again.

```{r R.options=list(tibble.width=78)}
glimpse(iris)
```



## Exercise
Begin by exploring the `iris` data with the `skim` function. What are all the things can you learn about the `Species` variable from this output? What are all the things you can learn about the `Sepal.Length` variable from this output?


### Solution


## Exercise
Now, look at all the pairwise relationships in the iris data set using `ggpairs`, using histograms on the diagonals for the numeric variables. Make sure you expand your plot window large enough to show all the data and axis labels clearly.

**A.** Does the relationship between Sepal Length and Species follow the same pattern as Sepal Width and Species?

**B.** Comment on the distribution of Petal Lengths. Which species do you think are included in the portion of the distribution with the higher center, and which species do you think are included in the portion of the distribution with the lower center?


### Solution


## Exercise
Consider the following data frame of summary statistics of the "setosa" species of iris:

```{r}
setosa_summary <- data.frame(
  Species = c("setosa", "setosa", "setosa", "setosa", "setosa", "setosa", "setosa", "setosa"),
  structure = c("Petal", "Sepal", "Petal", "Sepal", "Petal", "Sepal", "Petal", "Sepal"),
  dimension = c("Length_mean", "Length_mean", "Width_mean", "Width_mean", "Length_median", "Length_median", "Width_median", "Width_median"),
  value = c(1.462, 5.006, 0.246, 3.428, 1.5, 5, 0.2, 3.4),
  stringsAsFactors = FALSE
  )
setosa_summary
```

Right now, this data frame is not tidy: the value column mixes measures of multiple variables together (mean and median). Using functions from the `tidyr` package Put this data frame into a tidy layout. In other words, reshape the data frame so that the mean and median values for each plant structure and dimension combination are in separate columns. Make sure to print out your result.

### Solution

## Exercise
The iris data set itself is not arranged in a tidy layout: the four numeric columns combine information about what part of the plant (petal or sepal) is being measured with what dimension is being measured in the column names. In a tidy layout, a separate column should identify what plant structure was being measured for each observation, and the actual measurements of length and width should be in separate columns.

Put the iris data set into its correct "tidy" layout. When you think you have it, print out the first 5 rows of the data set for your solution.

<div class="attention">
**Hint**: Begin by adding an ID variable for each row of the iris data frame
</div>

### Solution
