---
title: "Data Types and Manipulation"
author: "Will Hopper"
date: "2/11/20"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
    css: ["default", "default-fonts", "../../assets/css/sds.css"]
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set("warning"=FALSE, "message"=FALSE, "fig.align" = 'center', "cache" = TRUE, dev='svg',prompt=FALSE)
```

## What is a data type?

A value's **data type** is defined by 

- the types of operations that can be performed with it in your software
- the range of values it can take on
- how the information is represented in hardware (e.g., CPU, memory)

R uses many different data types, but the most important to learn are: 
- **integer**
- **double**
- **character**, and
- **logical**

These are the **atomic** data types in R, because they are the smallest possible building blocks.

Many of R's data types have a loose correspondence to *statistical* variable types (e.g., numeric vs. categorical).

---

## Integers & Doubles 

Just as in mathematics, the **integer** data type can only represent whole numbers
  + e.g. 6

The **double** data type is used to represent all [real numbers](https://en.wikipedia.org/wiki/Real_number) (i.e., numbers with a fractional component).
  + e.g. 6.66666667
  + However, the **double** data type is not infinitely precise, and long decimal components are approximated by the computer.
  + R uses the double data type to represent numbers by default (i.e., 6 becomes 6.0000)

Together, doubles and integers are referred to as the **numeric** data types.

---

## Character


The **character** data type (sometimes referred to as the "string" data type) is used for representing textual data

To encode a value as character data in R, it must be wrapped in quotes (`" "` or `' '` are both acceptable)

```{r textdata_good}
a <- "foobar"
a
```

Without quotes, R will interpret the text as the name of an R object, and attempt to find that object and return its value. Missing quotes is a common source of "object not found" errors.

.errormsg[
```{r textdatabad, error=TRUE}
a <- foobar 
```
]
---

## Numbers as Numerics vs. Numbers as Characters

`4.2` can be represented in R as both a character and a double. But, only in one case can mathematical operations be performed on them.

.errormsg[
```{r badtyping, error=TRUE}
"4.2" + 1
```
]

```{r goodtyping}
4.2 + 1
```
This illustrates a practical differences between data types: some types do not support specific operations (e.g., characters do not support math operations)!

---

## Logical

Logical data can only take on 2 possible values: `TRUE` or `FALSE`

This type of datum is used to represent whether some state exists (is true) or does not exist (is false)

`TRUE` and `FALSE` *must* be upper-cased
  + *can* be abbreviated as `T` and `F`, but **I strongly recommend you do not do this**

This type of data is *extremely* useful for data manipulation.

---

## Checking Data Types 
You can use the `typeof` function to see what data type an object holds. 

```{r typeChecking}
typeof(2)
typeof(2L) # L at the end forces R to use integer, not double
typeof("herp")
typeof(FALSE)
```

---

## Coercion
Values stored in one data type can sometimes be changed to another data type. This transformation is called *coercion*

```{r coerce}
as.double("4.2") + 1
```

Sometimes when an operation requires a specific data type, R will coerce things to the proper types for us (called an implicit conversion).

```{r autocoerce}
TRUE + 1 # coerces TRUE to an integer 1 and adds
```

But things aren't always magically perfect

.errormsg[
```{r failcoerce, error=TRUE}
"TRUE" + 1
```
]

---

## Data Structures

Of course, in a language focused on analyzing data, it is helpful to have the ability to group multiple values together - this is where **data structures** come into plot.

Think of data structures in R as big containers for grouping together many values. After storing your data in these containers, you can reuse it multiple places (e.g. create an R object to store it) or access different subsets of it by position or name.

The four most used data structures, in order of flexibility, in R are:

1. Vectors
2. Matrices
3. Data Frames
4. Lists

In this course, we will be primarily working with vectors and data frames.

---

## Vectors
Vectors are 1 dimensional data structures which can hold numeric, logical, and character data.

Vectors are the most basic data structures in R. In fact, all the values we have worked with so far are represented internally by R as vectors of length 1!!

Data types may not be mixed in a vector (e.g. you cannot have some elements be characters and other be integers)

The individual values held in the vector are referred to as *elements*, and every vector has a *length* equal to the number of elements it contains. 

---

## Creating a vector

Vectors, no matter what type of data they hold, can be created by using the `c()` function in R, short for *concatenate*. 

Just place each value you want to be included in the vector inside the parenthesis, separated by a comma.

```{r vector_creation}
new_vector <- c(1, 10, 45, -1)
char_vector <- c("foo", "bar", "herp", "derp")
```

`c()` can combine existing vectors as well, not just create new ones.
```{r concatenate}
c(new_vector, c(1, 2, 3, 4, 5))
```

---

## Names 
After you create a vector, you can give the elements names, using the `names()` function, and a character vector.

```{r named_Vec}
new_vector
names(new_vector) <- c("A","B","C","D")
new_vector
```

This can be useful later when you want to pick out a few values from the larger vector (more about this later on).

---

## Vector Tricks
You can create a sequential numeric vector using the colon operator `:`, instead of typing out hundreds or thousands of values by hand.

```{r colon}
55:100
```

Sequences of other step sizes can be made with the `seq()` function
```{r seq}
seq(from = 5, to = 22, by = 3)
```

---

## Element-wise Math operations
You can apply the math operators we used in the first class to vectors as well.

When you add, divide, multiply, or subtract a set of vectors, R matches the vectors up by position and applies the operation to each pair of elements.

```{r math_vectors}
vec1 <- c(5, -1, 100, 75)
vec2 <- c(10, 3, 4, -4)
vec2 + vec1
vec2 * vec1
```

So, the first element of `vec1` gets added to the first element of `vec2`, etc...

If you want to find the grand sum or product of *all* the elements in a numeric vector, use the `sum()` and `prod()` functions, respectively.

---

## Recycling
If you just want to do something like multiply every number in a vector by 2, and your vector is 10,000 elements long, don't despair - you needn't type out 10,000 two's.

When two data structures have different lengths, R will **recycle** the input with the shorter length. In other words, it will automatically repeat the shorter input again and again until it matches the length of the longer one.

So if you have `vec1 <- c(5, -1, 100, 75)`, and you want to multiply every element by 2, you can just write `vec1 * 2` and R will automatically recycle the 2 into `vec1 * c(2, 2, 2, 2)` with no extra work on your part.

```{r recycling_vectors}
vec1 <- c(5, -1, 100, 75)
vec1 * 2
```

---

## Factors
Factors are a data structure that is a hybrid between a character vector, and a numeric vector.

Their purpose is to compactly represent the levels of a categorical variable.

The hybrid nature of a factor vector makes them tricky to work with, especially for new users of R.

The most important thing to remember is that the values you see when observing or "printing out" a factor vector **are not** the same as the values R stores and uses internally.

Let's take a look...

---

## Factors

Let's say we're representing data about smoker or non-smoker status using a 1 to indicate a smoker, and a 0 to indicate a non-smoker, like in the `ncbirths` data set.

And, since we know R cannot read our minds, and automagically know our zeros and ones represent categories instead of numbers, we choose to be good programmers and enter this data into R using the `factor` function.

```{r factors}
smoking <- factor(c(1, 0, 0, 0, 1, 0))
smoking
```

We see our 6 observations printed out, along with the `level` attribute, which tells us all the unique category levels in our data.

---

## Factors

But imagine that another researcher working on a similar project has coded their smoking data using 1 and 2, instead of 0 and 1, and we want to compare our data.

Instinct kicks in, and you decide just to add 1 to your data, so it matches the other researches data.

```{r factors1}
smoking
```

.errormsg[
```{r factors2, warning=TRUE}
smoking + 1
```
]

Oh right, addition and subtraction aren't meaningful for categories! So even though our data *looks like* numbers when printed, R remembers they're categories.

---

## Factors

But, we *really* know what we're doing™: we're not adding to do math, we're just re-coding our category labels the easy way.

Let's try coercing our factor vector back to a numeric variable, doing the addition there, and then using the `factor` function again.

```{r factor3}
smoking <- as.numeric(smoking)
smoking
```

Wait, how are they already one's and two's?????

---

## Factors
This weirdness is because behind the scenes, R was *already* using a 1 for non-smoker and a 2 for smoker.

The zeros and ones we saw printed out **were just labels** for the categories that R inferred based on our input data!


Printed Label | Internal Value
:-----------: | :------------:
     1        |       2       
     0        |       1       
     0        |       1       
     0        |       1       
     1        |       2       
     0        |       1       

---

## Factors
R uses numbers as the internal representation for your categories even when the input data are explicitly character data.

```{r factor4}
smoker <- factor(c("smoker","nonsmoker","nonsmoker",
                   "nonsmoker", "smoker","nonsmoker")
                 )
smoker
as.numeric(smoker)
```

---

## Factors
If you really do need to use addition on your numerically-labeled factor vector, the easiest thing to do is convert it to a character vector with `as.character` (which will use the labels) and *then* use `as.numeric`
```{r factor5}
smoker <- factor(c(1, 0, 0, 0, 1, 0))
as.character(smoker)
as.numeric(as.character(smoker))
as.numeric(as.character(smoker)) + 10
```

---

## Factors
The best thing is to not use numeric labels for your categories, so you avoid this confusion.

If your input data are numeric, you can give each level as non-numeric label using the `labels` argument to the `factor` function.

```{r factor6}
smoker <- factor(c(1, 0, 0, 0, 1, 0),
                 labels = c("nonsmoker", "smoker")
                 )
smoker
```

Just make sure you supply the labels in the correct order!! The value 0 will be the first level, and 1 the second level, so we supply the labels with nonsmoker first, then smoker second, to preserve the interpretation of our coding.

---

## Data Frames

Data Frames are a 2D, rectangular data structure that allow for different data types to be stored in each column.

A Data Frame is usually the best way to store and work with data that mixes qualitative and quantitative variables.

And don't forget to keep your data frames tidy!!

---

## Creating Data Frames
Data frames can be created by passing `name = value` pairs to the `data.frame()` function.

The values should be vectors (of any type) and the names should be unquoted strings of text, which will be used to label the column.


Importantly, all the vectors stored in the data frame must be of *identical length*.

```{r dataframeexample}
df <- data.frame(x =     c(  1,    4,   4,     2  ),
                 y =     c(  3,    3,   1,     4  ),
                 month = c("Sep","Oct","Nov","Jan"),
                 stringsAsFactors = FALSE)
df
```

---

## stringsAsFactors?
This example also shows an extra argument, `stringsAsFactors = FALSE`, that was **not** a variable to store in the data frame. This argument controls how R interprets character vectors when forming the data frame.

Under the default setting, where `stringsAsFactors` is `TRUE`, R converts all character vectors into **factor** vectors inside the data frame. If `stringsAsFactors = FALSE`, then R leaves the data alone when forming the data frame.

Since there are lots of confusing "gotchas" to worry about with factors (even more than ones I showed you earlier!), my advice is to override the default, and use `stringsAsFactors = FALSE`.

Note that **a lot** of data import related functions have this argument, so get into the habit of specifying `stringsAsFactors = FALSE` when you use functions that create data frames (e.g., `data.frame()`, `read.csv()`, `read.table()`, etc.)

If you want to know more about the reasons for all this nonsense, [stringsAsFactors: An unauthorized biography](https://simplystatistics.org/2015/07/24/stringsasfactors-an-unauthorized-biography/) and [stringsAsFactors = sigh](http://notstatschat.tumblr.com/post/124987394001/stringsasfactors-sigh) are two great reads.

---

## Data Frame Tricks (and Traps)
The `data.frame()` function can also recycle input, so you can save some keystrokes if you have a column that has only a few unique values that repeat.

R compares the length of the first vector to the length of each vector after it. If the lengths don't match, R will check if the length of the longer vector is a multiple of the length of the shorter vector.

If so, R will repeat the shorter vector over and over so the lengths match up. In this example, the `x` variable has length 4, and the `month` vector has length 1; since 4 is a multiple of 1 (1 x 4 = 4), R will repeat the `month` vector 4 times so the lengths match.

```{r goodrecycle}
data.frame(x = c(1, 4, 4, 2),
           month = "Sep"
           )
```

---

## Data Frame Tricks (and Traps)
The `data.frame()` function will throw an error if the length of the longer vector is not a multiple of the shorter vector's lengths.

Here, the `x` variable still has length 4, but the `month` vector has length 3; since 4 is not multiple of 3, an error occurs.

.errormsg[
```{r badrecycle, error=TRUE}
data.frame(x = c(1, 4, 4, 2),
           month=c("Sep", "Oct", "Nov")
           ) 
```
]

---

## Recycling Advice
Recycling enables concise code, but can make for hard to spot bugs when you didn't intended for recycling to happen. 

This is especially likely when the number of columns in your data frame starts to grow - when you have 10 columns, what are the odds the recycling works the way you want for *all* the columns? 

My advice: Favor being explicit about your vector lengths when programming, and only rely on length 1 inputs being recycled correctly!

---
