
---
title: 'Problem Set 07: Multiple Regression'
author: "Will Hopper"
output:
  html_document:
    css: "https://wjhopper.github.io/sds201_labs/assets/css/lab.css"
    theme: lumen
    toc: yes
    toc_float:
      collapsed: false
    toc_depth: 2
    df_print: paged
    self_contained: true
editor_options: 
  chunk_output_type: console
---

In this problem set we will continue our work with the [FiveThirtyEight hate crimes data set](https://fivethirtyeight.com/features/higher-rates-of-hate-crimes-are-tied-to-income-inequality/). This time, we'll model these data using multiple explanatory variable within the same model, and try out different model architectures (i.e., parallel slopes and interaction models).

First load the necessary packages:

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(moderndive)
```

Next, load the data set from the internet:

```{r}
hate_crimes <- read.csv("https://wjhopper.github.io/sds201_labs/data/hate_crimes.csv", stringsAsFactors = FALSE)
```

Just to remind you, the data set has 51 observations from the 11 following variables:

- `state`: State name, includes District of Columbia (i.e., Washington D.C.)
- `percent_hs_degree`: Proportion of adults 25 and older with a high-school degree (2009)
- `share_non_citizen`: Proportion of the population that are not U.S. citizens
- `share_white_poverty`: Proportion of white residents who are living in poverty
- `gini_index`: GINI inequality index (higher = greater inequality)
- `share_non_white`: Proportion of the population that is not white
- `percent_trump`: Proportion of 2016 U.S. presidential voters who voted for Donald Trump
- `hate_crime_rate`: Hate crimes per 100,000 people between Nov. 9 & Nov. 18, 2016 according to Southern Poverty Law Center 
- `median_house_inc`: Median household income, divided into "high" and "low" categories
- `unemployment`: Unemployment rate divided into "high" and "low" categories
- `urbanization`: Proportion of the population living in metropolitan areas, divided into high" and "low" categories


Be sure to explore the data using the `skim` or `glimpse` function, or the RStudio data viewer (just remember, don't print out the whole data set in your submission!).

# An Additive Regression Model

For the exercises in this section, model the relationship between the hate crime rate (`hate_crime_rate`) as the **outcome**, and both the GINI inequality index (`gini_index`) and the median household income category (`median_house_inc`) as  **explanatory** variables, using a **parallel slopes** multiple regression model.

## Exercise
Visualize the relationship between the hate crime rate, the GINI inequality index, and the median household income category. Your visualization should include regression lines that represent the predictions of the **parallel slopes** multiple regression model. Describe the relationship between these three variables.

### Solution


## Exercise
Fit the parallel slopes regression model visualized in the previous question, and report the regression table.

### Solution.


## Exercise
Based on this regression model, what is:

1. The value of the $b_0$ parameter?
2. The value of the $b_1$ parameter?
3. The value of the $b_2$ parameter?
4. The y intercept of the baseline income group?
5. The y intercept of the "high" income group?
6. The difference between the y intercept of the "low" income group, and the "high" income group?
7. The difference between hate crime rate of the "low" income group and the "high" income group when the GINI index is .44?
8. The predicted increase in hate crime rate for a one percentage point increase in the GINI index for a "high" income state?
9. The predicted increase in hate crime rate for a one percentage point increase in the GINI index for a "low" income state?

*HINT:* Some of the items in this list are identical. In other words, some of these items refer to the same quantities in the regression table.

### Solution


## Exercise
Based on this model, what hate crime rate would you expect to occur in a "high" income state with a GINI index of .5? What hate crime rate would you expect to occur in a "low" income state with a GINI index of .5?

**HINT**: You may find it helpful to write out the regression model in GLM notation as a first step.

### Solution


## Exercise
The state of Colorado belonged to the "high" median income group, had a GINI index of 0.457 in 2016, and a hate crime rate of .39 per 100,000 people. 

Back in Exercise 5 of Problem Set 6, we found that when we predicted the hate crime rate in Colorado with just the GINI index alone, our bivariate regression model had a residual error of 0.088. How much more accurate is this multiple regression model for predicting the hate crime rate in Colorado?

### Solution


# An Interactive Regression Model
For the exercises in this section, model the relationship between the hate crime rate (`hate_crime_rate`) as the **outcome**, and both the GINI inequality index (`gini_index`) and the median household income category (`median_house_inc`) as  **explanatory** variables, using a multiple regression model that allows for an **interaction**.

## Exercise
Visualize the relationship between the hate crime rate, the GINI inequality index, and the median household income category. Your visualization should include regression lines that represent the predictions of the **interactive** multiple regression model. What, if anything, important has changed from the parallel slopes model?

### Solution


## Exercise
Fit the interaction model visualized in the previous question, and report the regression table.

### Solution.


## Exercise
Based on this regression model, what is:

1. The value of the $b_0$ parameter?
2. The value of the $b_1$ parameter?
3. The value of the $b_2$ parameter?
4. The value of the $b_3$ parameter?
5. The predicted hate crime rate for the baseline group when the GINI index is 0?
6. The y intercept of the "low" income group?
7. The predicted increase in hate crime rate for a one percentage point increase in the GINI index for a "high" income state?
8. The predicted increase in hate crime rate for a one percentage point increase in the GINI index for a "low" income state?

### Solution.


## Exercise
Based on this model, what hate crime rate would you expect to occur in a "low" income state with a GINI index of .5?

**HINT**: You may find it helpful to write out the regression model in GLM notation as a first step.

### Solution


# Model Comparison

## Exercise
Compute the proportion of variance in hate crimes accounted for by both the parallel slopes model, and by the interaction model. Based on these values, and the visualizations you created in Exercises 2 and 7, which model do you prefer: the parallel slopes model, or the interaction model. Why?

### Solution

