
---
title: "Reshaping and Exploring Data"
author: "Will Hopper"
date: "2/11/20"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
    css: ["default", "default-fonts", "../../assets/css/sds.css", "../../assets/css/tidyCellHighlighting.css"]
editor_options:
  chunk_output_type: console
---

```{r setup,echo=FALSE,message=FALSE,warning=FALSE}
library(tidyr)
library(dplyr)
library(GGally)
library(skimr)
library(whoppeR)
knitr::opts_chunk$set(warning=FALSE, message=FALSE, cache = TRUE, dev='svg', fig.align = 'center')
options(knitr.table.format = "html")
```

## Real data's dirty secret
As you search for project data, you might experience a moment of profound dissapointment:

1. You come up with a cool research question
2. You actually find a source of relevant data
3. You open up the data, and... wait what?

The data is so awkward and messy, you've got tons of work to do before you get down the business of statistical inference.

Today, we're going to be talking about how to deal with one of the most common kinds of messy data: data that is not in a "tidy" layout.

---

## Reshaping with `tidyr`
The tools from the `tidyr` package help you re-arrange data frames, with an emphasis on helping you *normalize* datasets.

There are 4 main functions for re-arranging data in the `tidyr` package:

1. `gather()`: transforms data from wide(r) to long(er)
2. `spread()`: transforms data from long(er) to wide(r)
3. `separate()`: splits a single column into multiple columns
4. `unite()`: combines multiple columns into a single column

---

## The temperature data
We'll demonstrate how to use each function with a small dataset, so its easy to see what changes get made. So, install `tidyr`, and load the [temperature dataset](../data/temperature.csv).

```{r, eval=FALSE, cache=FALSE}
install.packages("tidyr")
library(tidyr)
temperature <- read.csv("https://wjhopper.github.io/psych640-labs/data/temperature.csv")
temperature
```

```{r import_temp_data, echo=FALSE, cache=FALSE}
temperature <- read.csv("../../data/temperature.csv")
tmp_initial <- temperature
```

```{r display_temp_data, echo=FALSE}
knitr::kable(temperature, caption = "High Temperatures (Farenheight)", format = "html")
```

---

## The temperature data
```{r, ref.label="display_temp_data", echo=FALSE}
knitr::kable(temperature, caption = "High Temperatures (Farenheight)", format = "html")
```

The `temperature` dataset has measurements of 3 variables: **City**, **Date**, and **Temperature**.

But the layout of the dataset is in so-called "wide form", because values of the **Date** variable (Feb_19_2017, Feb_18_2017, and Feb_17_2017) are used as unique variables themselves.

Instead, the dates *should* be in the cells of the table as observed values of the **Date** variable.

---

## The `gather()` function 
Let's use the `gather` function to move the dates in the column headers down into the table under a new variable **date**, and move the actual temperature measurements into a new variable called **temp**.

`gather` takes 4 required arguments:

1. `data`: the name of the data frame to re-arrange
2. `key`: the name of the to-be-created variable that will hold values currently in use as variables
3. `value`: the name of the to-be-created variable that will hold the values from the variables to be gathered.
4. `...`: A bare list of columns to gather (or columns *not* to gather)

---

## The `gather()` function 
Let's make this more specific to our situation.

```{r,eval=FALSE}
gather(data = temperature,
       key = "date",
       value = "temp",
       -city)
```

- The `key` column will be named "date", because it will hold the dates that were previously used as individual variables
- The `value` column will be named "temp" because the values in each of the Feb_19_2017, Feb_18_2017, and Feb_17_2017 variables were our actual temperature measurements.
- We want to gather values from all the columns *except* the city column, so we can write `-city`. Alternatively, we could write `Feb_19_2017, Feb_18_2017, and Feb_17_2017`.

---

class: gather, spthis

## The `gather()` function 
So this:
```r
temperature <- gather(data = temperature, key = "date", value = "temp", -city)
```
.pull-left[
turns this:
```{r, echo=FALSE}
knitr::kable(temperature)
```
]

.pull-right[
into this:

```{r, echo=FALSE}
temperature <- gather(data = temperature, key = "date", value = "temp", -city)
knitr::kable(temperature)
```
]

---

class: gather2

## The `gather()` function
Critically, the temperature observations are still matched up with the same date!

```{r, echo=FALSE}
knitr::kable(tmp_initial) 
```

```{r, echo=FALSE}
knitr::kable(temperature)
```

---

## The `spread()` function
The `spread` function is the complement to the `gather` function, so we can use it to undo this transformation. 

Its job is to takes values from an existing variable in the dataset and "spread" them out, creating a new variable for each unique value in the old variable.

The `spread` function takes 3 required arguments:

1. `data`: the name of the data frame to re-arrange
2. `key`: the name of the existing variable whose *N* unique values will be "spread out" into *N* new variables
3. `value`: the name of the existing variable whose values will be used as values in the *N* to-be-created variables.

---

## The `spread()` function
So to undo the wider-to-longer normalization we just performed, we would use this function call:

```{r,eval=FALSE}
spread(data = temperature,
       key = date,
       value = temp)
```

- the `key` argument is the "date" column, because the "date" column holds the values we want to be new variables
- the `value` argument is the "temp" column, because we want the temperature measurements to be observations in the 3 date variables

We don't have to specify the names of the new variable we're creating, because they are going to come from the values in the "date" variable.

---

class: sothis

## The `spread()` function
So this:
```r
spread(data = temperature, key = date, value = temp)
```

.pull-left[
turns this:
```{r, echo=FALSE}
knitr::kable(temperature)
```
]

.pull-right[
back into this:
```{r, echo=FALSE}
knitr::kable(spread(data = temperature, key = date, value = temp))
```
]

---

## Common `spread()` problems
A common issue encountered with the `spread` function is the somewhat mysterious error "<code style="color: red;">Each row of output must be identified by a unique combination of keys.</code>".

Basically, this means is that when the columns given as the `key` and `value` arguments are removed, the rows of your data frame are not unique. 

Thus, `spread` can't figure out which observations from the `value` column belong in which rows when it reshapes that data. Rather than potentially corrupt your data, it gives up instead.

In general, the solution is to give each row a unique ID number like so:

```{r eval = FALSE} 
temperature$id <- 1:nrow(temperature)
```

---

## Common `spread()` problems
For example, I can't spread the `X` variable in this data frame into new `A` and `B` variables.
```{r notunique,echo=FALSE}
set.seed(10)
d <- data.frame(X = c("A","A","B","B"),
                Y = c(1, 1, 2, 2),
                Z = sample(10:200, 4))
knitr::kable(d)
```

.errormsg[
```{r trytospread, error=TRUE}
spread(d, key=X, value=Z)
```
]

---

## Common `spread()` problems
To see why, imagine that the `X` and `Z` columns were gone. With just the `Y` column left, rows 1 and 2 are duplicates, as are 3 and 4.
```{r notunique2,echo=FALSE}
knitr::kable(d)
```

Without unique rows to match against, R can't tell whether 146 or 82 should go in the first row of column `A`. It *could guess* based on row order, but it won't.

---

class: spread_table

## Common `spread()` problems
If we add an ID variable, we can reshape the data.
.pull-left[
```{r adduniqueID1, echo=1}
d$id <- 1:4
knitr::kable(d)
```
]

.pull-right[
```{r show_spread, eval=FALSE}
spread(d, key=X, value=Z)
```
```{r actuallyspread, echo=FALSE}
knitr::kable(spread(d, key=X, value=Z)) 
```
]

This may seem like an odd result, but if you look carefully, everything is correctly matched up. It's just that "B"'s were ever seen with a 1 in `Y`. So, all the rows associated with Y=1 are missing in column B. The same pattern holds for column A and Y=2.

---

## The `separate()` function
The `separate` function is used to transform one variable in your dataset into many variables in your dataset.
This is useful when multiple variables have been concatenated into a single variable, and you would like to examine these variables separately.

For example, in our normalized `temperature` dataset, the `date` variable is actually a combination of a month, day of month, and year variable, concatenated together.

```{r, echo=FALSE}
knitr::kable(temperature)
```

---

## The `separate()` function
Let's separate this amalgamated "date" variable into 3 different variables, one for the month, day, and year.

The `separate` function has 3 required arguments

1. `data`: the name of the data frame to re-arrange
2. `col`: the bare name of the variable to be split-up
3. `into`: a character vector of names for the to-be-created variables

`separate` will split the values in the `col` variable on any "punctuation" character (e.g., `_`,  `-`, `.`, `:`, etc.).

---

## The `separate()` function
In our case, we would use the separate function like so:
```{r separate, eval=FALSE}
separate(data = temperature,
         col = date,
         into = c("month", "day", "year")
         )
```

- We want to separate the `col`umn named `date` *into* 3 new variables named "month", "day", and "year"

--

- We are going to find the values that should go into the month, day, and year variables by splitting up the values in the `date` variable at every underscore character.

--

- So all the values like `"Feb_19_2017"` are going to turn into `"Feb"`, `"19"` and `"2017"`

---

class: sothis
## The `separate()` function
So this:
```r
temperature <- separate(temperature, col=date, into=c("month", "day", "year"))
```

.pull-left[
turns this:
```{r, echo=FALSE}
knitr::kable(prettifyTable(temperature, rows = 3))
```
]

.pull-right[
into this:
```{r, echo=FALSE}
temperature <- separate(temperature, col=date, into=c("month", "day", "year"), sep = "_")
knitr::kable(prettifyTable(temperature, rows = 3))
```
]

---


## The `unite()` function
The `unite` function is the complement to the separate function. As you can imagine, its job is to combine multiple variables into a single variable. 

The `unite` function has 3 required arguments, and 2 other arguments you often want to specify:

1. `data`: the name of the data frame to re-arrange
2. `col`: the bare name of the to-be-created amalgam variable
3. `...` : bare columns whose values are to be combined row-wise and placed inside the new variable `col`
4. `sep`: the separator to use between combined values (defaults to `"_"`).
5. `remove`: a logical value indicating whether or not to remove columns from the data frame after combining them (defaults to `TRUE`).

---

class: sothis

## The `unite()` function
So to undo our separation of the `date` variable, we could use this:
```r
unite(temperature, col=date, month, day, year)
```

.pull-left[
to turn this:
```{r, echo=FALSE}
knitr::kable(prettifyTable(temperature, rows = 3))
```
]

.pull-right[
back into this:
```{r, echo=FALSE}
knitr::kable(prettifyTable(unite(temperature,
                           col=date,
                           month, day, year),
                     rows = 3))
```
]

---


## Resources & More
For more information about the *theory* behind the `tidyr` package, check out the [package vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) and the [tidy data paper](http://vita.had.co.nz/papers/tidy-data.html).

You may have noticed that the API (i.e., standard syntax) for `tidyr` functions is not unlike that of the functions in `dplyr`. This is not a coincidence; both packages have the same core authors and are part of the "[tidyverse](https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/)" collection of R packages.

However, `tidyr` is *not* the only choice for reshaping data. The `reshape2` package is also quite useful. Its focus is on reshaping more broadly (e.g., from one type of data structure to another, reshaping through aggregation, etc.). If you need a more general reshaping tool, start with the following `reshape2` tutorials:

- [My reshape2 lab](reshape.html)
- [A brief but good intro to reshape2](http://seananderson.ca/2013/10/19/reshape.html)
- [Hadley Wickham's intro to reshape2](http://had.co.nz/reshape/introduction.pdf)

---

## Practice

Consider the following data set:

```{r face_rt}
wide <- data.frame(ID = 1:6,
                   Face.1 = c(411,723,325,456,579,612),
                   Face.2 = c(123,300,400,500,600,654),
                   Face.3 = c(1457,1000,569,896,956,2345)
                   )
```

```{r face_rt_show, echo=FALSE}
knitr::kable(wide)
```


---

## Practice

Use the `gather` and `separate` functions to put this data set into tidy form, and then use `unite` and `spread` to put in back in the original wide format.

```{r face_rt_show2, echo=FALSE}
knitr::kable(wide)
```

---

class: inverse, middle, center

# Exploring Data

---

## Exploring New Datasets
The next two things we're going to talk about are tools for digging into a new data set, and finding out what the heck is in there. They are:

1. The `skim` function from the `skimr` package
  ```{r, eval=FALSE}
  install.packages('skimr')
  library(skimr)
  ```
2. The `ggpairs` function from the `GGally` package
  ```{r, eval=FALSE}
  install.packages('GGally')
  library(GGally)
  ```

Both of these are most useful in the context of a larger data set, so let's bring in the `ncbirths` data set as usual.

```{r, eval=FALSE}
ncbirths <- read.csv("https://wjhopper.github.io/SDS-201/data/ncbirths.csv",
                     stringsAsFactors = FALSE)
```

```{r echo=FALSE}
ncbirths <- read.csv("../../data/ncbirths.csv", stringsAsFactors = FALSE)
```

---

class: skimmed

## skim
```{r, results='asis'}
skim(ncbirths)
```

---

class: skimmed
## skim lite
Sometimes, the characters used to draw the histograms can cause problems. If you find yourself encoutering strange unicode errors about character  <U+2587> , try using the `skim_without_charts()` function instead of `skim`:

```{r, results='asis'}
skim_without_charts(ncbirths, weight, gained) # just these two to demonstrate
```

---

## ggpairs

`ggpairs` is a function to visualize the relationship between all possible pairs of variables in your data set, in what is often called a "scatter plot matrix".

If you have 2 variables, your plot will be a 2x2 matrix with 4 sub-plots. If you have 3, a 3x3 matrix with 9 sub-plots. If you have 4, a 4x3 matrix with 16 sub-plots. You get the idea...

Why is this important? Because you run out of room on a laptop screen fast.

---

## ggpairs

```{r, message=FALSE}
ggpairs(ncbirths)
```

---

Let's just look at the first 4 variables to get a better idea of what's happening
```{r first_four_pairs, message=FALSE, fig.height=5.25, fig.width=7}
ggpairs(ncbirths, columns = 1:4)
```

---

Mostly same information in different forms above and below the diagonal

```{r, ref.label="first_four_pairs", message=FALSE, fig.height=5.25, fig.width=7}
```

---

If you want histograms instead of smoothed density estimates along the diagonal, use this:
```{r hist_diagonal, message=FALSE, fig.height=5.25, fig.width=7}
ggpairs(ncbirths, columns = 1:4, diag = list(continuous='barDiag'))
```
