
---
title: "Data Types and Manipulation"
author: "Will Hopper"
date: "2/11/20"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
    css: ["default", "default-fonts", "../../assets/css/sds.css"]
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set("warning"=FALSE, "message"=FALSE, "fig.align" = 'center', "cache" = TRUE, dev='svg',prompt=FALSE)
```

## What is a data type?

A value's **data type** is defined by 

- the types of operations that can be performed with it in your software
- the range of values it can take on
- how the information is represented in hardware (e.g., CPU, memory)

R uses many different data types, but the most important to learn are: 
- **integer**
- **double**
- **character**, and
- **logical**

These are the **atomic** data types in R, because they are the smallest possible building blocks.

Many of R's data types have a loose correspondence to *statistical* variable types (e.g., numeric vs. categorical).

---

## Integers & Doubles 

Just as in mathematics, the **integer** data type can only represent whole numbers
  + e.g. 6

The **double** data type is used to represent all [real numbers](https://en.wikipedia.org/wiki/Real_number) (i.e., numbers with a fractional component).
  + e.g. 6.66666667
  + However, the **double** data type is not infinitely precise, and long decimal components are approximated by the computer.
  + R uses the double data type to represent numbers by default (i.e., 6 becomes 6.0000)

Together, doubles and integers are referred to as the **numeric** data types.

---

## Character


The **character** data type (sometimes referred to as the "string" data type) is used for representing textual data

To encode a value as character data in R, it must be wrapped in quotes (`" "` or `' '` are both acceptable)

```{r textdata_good}
a <- "foobar"
a
```

Without quotes, R will interpret the text as the name of an R object, and attempt to find that object and return its value. Missing quotes is a common source of "object not found" errors.

.errormsg[
```{r textdatabad, error=TRUE}
a <- foobar 
```
]

---

## Numbers as Numerics vs. Numbers as Characters

`4.2` can be represented in R as both a character and a double. But, only in one case can mathematical operations be performed on them.

.errormsg[
```{r badtyping, error=TRUE}
"4.2" + 1
```
]

```{r goodtyping}
4.2 + 1
```
This illustrates a practical differences between data types: some types do not support specific operations (e.g., characters do not support math operations)!

---

## Logical

Logical data can only take on 2 possible values: `TRUE` or `FALSE`

This type of datum is used to represent whether some state exists (is true) or does not exist (is false)

`TRUE` and `FALSE` *must* be upper-cased
  + *can* be abbreviated as `T` and `F`, but **I strongly recommend you do not do this**

This type of data is *extremely* useful for data manipulation.

---

## Checking Data Types 
You can use the `typeof` function to see what data type an object holds. 

```{r typeChecking}
typeof(2)
typeof(2L) # L at the end forces R to use integer, not double
typeof("herp")
typeof(FALSE)
```

---

## Coercion
Values stored in one data type can sometimes be changed to another data type. This transformation is called *coercion*

```{r coerce}
as.double("4.2") + 1
```

Sometimes when an operation requires a specific data type, R will coerce things to the proper types for us (called an implicit conversion).

```{r autocoerce}
TRUE + 1 # coerces TRUE to an integer 1 and adds
```

But things aren't always magically perfect

.errormsg[
```{r failcoerce, error=TRUE}
"TRUE" + 1
```
]

---

## Data Structures

Of course, in a language focused on analyzing data, it is helpful to have the ability to group multiple values together - this is where **data structures** come into play.

Think of data structures in R as big containers for grouping together many values. After storing your data in these containers, you can reuse it multiple places (e.g. create an R object to store it) or access different subsets of it by position or name.

The four most used data structures, in order of flexibility, in R are:

1. Vectors
2. Matrices
3. Data Frames
4. Lists

In this course, we will be primarily working with vectors and data frames.

---

## Vectors
Vectors are 1 dimensional data structures which can hold numeric, logical, and character data.

Vectors are the most basic data structures in R. In fact, all the values we have worked with so far are represented internally by R as vectors of length 1!!

Data types may not be mixed in a vector (e.g. you cannot have some elements be characters and other be integers)

The individual values held in the vector are referred to as *elements*, and every vector has a *length* equal to the number of elements it contains. 

---

## Creating a vector

Vectors, no matter what type of data they hold, can be created by using the `c()` function in R, short for *concatenate*. 

Just place each value you want to be included in the vector inside the parenthesis, separated by a comma.

```{r vector_creation}
new_vector <- c(1, 10, 45, -1)
char_vector <- c("foo", "bar", "herp", "derp")
```

`c()` can combine existing vectors as well, not just create new ones.
```{r concatenate}
c(new_vector, c(1, 2, 3, 4, 5))
```

---

## Names 
After you create a vector, you can give the elements names, using the `names()` function, and a character vector.

```{r named_Vec}
new_vector
names(new_vector) <- c("A","B","C","D")
new_vector
```

This can be useful later when you want to pick out a few values from the larger vector (more about this later on).

---

## Vector Tricks
You can create a sequential numeric vector using the colon operator `:`, instead of typing out hundreds or thousands of values by hand.

```{r colon}
55:100
```

Sequences of other step sizes can be made with the `seq()` function
```{r seq}
seq(from = 5, to = 22, by = 3)
```

---

## Element-wise Math operations
You can apply the math operators we used in the first class to vectors as well.

When you add, divide, multiply, or subtract a set of vectors, R matches the vectors up by position and applies the operation to each pair of elements.

```{r math_vectors}
vec1 <- c(5, -1, 100, 75)
vec2 <- c(10, 3, 4, -4)
vec2 + vec1
vec2 * vec1
```

So, the first element of `vec1` gets added to the first element of `vec2`, etc...

If you want to find the grand sum or product of *all* the elements in a numeric vector, use the `sum()` and `prod()` functions, respectively.

---

## Recycling
If you just want to do something like multiply every number in a vector by 2, and your vector is 10,000 elements long, don't despair - you needn't type out 10,000 two's.

When two data structures have different lengths, R will **recycle** the input with the shorter length. In other words, it will automatically repeat the shorter input again and again until it matches the length of the longer one.

So if you have `vec1 <- c(5, -1, 100, 75)`, and you want to multiply every element by 2, you can just write `vec1 * 2` and R will automatically recycle the 2 into `vec1 * c(2, 2, 2, 2)` with no extra work on your part.

```{r recycling_vectors}
vec1 <- c(5, -1, 100, 75)
vec1 * 2
```

---

## Factors
Factors are a data structure that is a hybrid between a character vector, and a numeric vector.

Their purpose is to compactly represent the levels of a categorical variable.

The hybrid nature of a factor vector makes them tricky to work with, especially for new users of R.

The most important thing to remember is that the values you see when observing or "printing out" a factor vector **are not** the same as the values R stores and uses internally.

Let's take a look...

---

## Factors

Let's say we're representing data about smoker or non-smoker status using a 1 to indicate a smoker, and a 0 to indicate a non-smoker, like in the `ncbirths` data set.

And, since we know R cannot read our minds, and automagically know our zeros and ones represent categories instead of numbers, we choose to be good programmers and enter this data into R using the `factor` function.

```{r factors}
smoking <- factor(c(1, 0, 0, 0, 1, 0))
smoking
```

We see our 6 observations printed out, along with the `level` attribute, which tells us all the unique category levels in our data.

---

## Factors

But imagine that another researcher working on a similar project has coded their smoking data using 1 and 2, instead of 0 and 1, and we want to compare our data.

Instinct kicks in, and you decide just to add 1 to your data, so it matches the other researches data.

```{r factors1}
smoking
```

.errormsg[
```{r factors2, warning=TRUE}
smoking + 1
```
]

Oh right, addition and subtraction aren't meaningful for categories! So even though our data *looks like* numbers when printed, R remembers they're categories.

---

## Factors

But, we *really* know what we're doing™: we're not adding to do math, we're just re-coding our category labels the easy way.

Let's try coercing our factor vector back to a numeric variable, doing the addition there, and then using the `factor` function again.

```{r factor3}
smoking <- as.numeric(smoking)
smoking
```

Wait, how are they already one's and two's?????

---

## Factors
This weirdness is because behind the scenes, R was *already* using a 1 for non-smoker and a 2 for smoker.

The zeros and ones we saw printed out **were just labels** for the categories that R inferred based on our input data!


Printed Label | Internal Value
:-----------: | :------------:
     1        |       2       
     0        |       1       
     0        |       1       
     0        |       1       
     1        |       2       
     0        |       1       

---

## Factors
R uses numbers as the internal representation for your categories even when the input data are explicitly character data.

```{r factor4}
smoker <- factor(c("smoker","nonsmoker","nonsmoker",
                   "nonsmoker", "smoker","nonsmoker")
                 )
smoker
as.numeric(smoker)
```

---

## Factors
If you really do need to use addition on your numerically-labeled factor vector, the easiest thing to do is convert it to a character vector with `as.character` (which will use the labels) and *then* use `as.numeric`
```{r factor5}
smoker <- factor(c(1, 0, 0, 0, 1, 0))
as.character(smoker)
as.numeric(as.character(smoker))
as.numeric(as.character(smoker)) + 10
```

---

## Factors
The best thing is to not use numeric labels for your categories, so you avoid this confusion.

If your input data are numeric, you can give each level as non-numeric label using the `labels` argument to the `factor` function.

```{r factor6}
smoker <- factor(c(1, 0, 0, 0, 1, 0),
                 labels = c("nonsmoker", "smoker")
                 )
smoker
```

Just make sure you supply the labels in the correct order!! The value 0 will be the first level, and 1 the second level, so we supply the labels with nonsmoker first, then smoker second, to preserve the interpretation of our coding.

---

## Data Frames

Data Frames are a 2D, rectangular data structure that allow for different data types to be stored in each column.

A Data Frame is usually the best way to store and work with data that mixes qualitative and quantitative variables.

And don't forget to keep your data frames tidy!!

---

## Creating Data Frames
Data frames can be created by passing `name = value` pairs to the `data.frame()` function.

The values should be vectors (of any type) and the names should be unquoted strings of text, which will be used to label the column.


Importantly, all the vectors stored in the data frame must be of *identical length*.

```{r dataframeexample}
finances <- data.frame(sales = c(100,   150,  400,  350 ),
                       costs = c(300,   300,  100,  200 ),
                       month = c("Sep","Oct","Nov","Jan"),
                       stringsAsFactors = FALSE)
finances
```

---

## stringsAsFactors?
This example also shows an extra argument, `stringsAsFactors = FALSE`, that was **not** a variable to store in the data frame. This argument controls how R interprets character vectors when forming the data frame.

Under the default setting, where `stringsAsFactors` is `TRUE`, R converts all character vectors into **factor** vectors inside the data frame. If `stringsAsFactors = FALSE`, then R leaves the data alone when forming the data frame.

Since there are lots of confusing "gotchas" to worry about with factors (even more than ones I showed you earlier!), my advice is to override the default, and use `stringsAsFactors = FALSE`.

Note that **a lot** of data import related functions have this argument, so get into the habit of specifying `stringsAsFactors = FALSE` when you use functions that create data frames (e.g., `data.frame()`, `read.csv()`, `read.table()`, etc.)

If you want to know more about the reasons for all this nonsense, [stringsAsFactors: An unauthorized biography](https://simplystatistics.org/2015/07/24/stringsasfactors-an-unauthorized-biography/) and [stringsAsFactors = sigh](http://notstatschat.tumblr.com/post/124987394001/stringsasfactors-sigh) are two great reads.

---

## Data Frame Tricks (and Traps)
The `data.frame()` function can also recycle input, so you can save some keystrokes if you have a column that has only a few unique values that repeat.

R compares the length of the first vector to the length of each vector after it. If the lengths don't match, R will check if the length of the longer vector is a multiple of the length of the shorter vector.

If so, R will repeat the shorter vector over and over so the lengths match up. In this example, the `sales` variable has length 4, and the `month` vector has length 1; since 4 is a multiple of 1 (1 x 4 = 4), R will repeat the `month` vector 4 times so the lengths match.

```{r goodrecycle}
data.frame(sales = c(100, 400, 400, 350),
           month = "Sep"
           )
```

---

## Data Frame Tricks (and Traps)
The `data.frame()` function will throw an error if the length of the longer vector is not a multiple of the shorter vector's lengths.

Here, the `sales` variable still has length 4, but the `month` vector has length 3; since 4 is not multiple of 3, an error occurs.

.errormsg[
```{r badrecycle, error=TRUE}
data.frame(sales = c(100, 400, 400, 350),
           month = c("Sep", "Oct", "Nov")
           ) 
```
]

---

## Recycling Advice
Recycling enables concise code, but can make for hard to spot bugs when you didn't intended for recycling to happen. 

This is especially likely when the number of columns in your data frame starts to grow - when you have 10 columns, what are the odds the recycling works the way you want for *all* the columns? 

My advice: Favor being explicit about your vector lengths when programming, and only rely on length 1 inputs being recycled correctly!

---

class: inverse, middle, center

# Indexing

---
## Indexing

All throughout your data analysis pipeline, you will face the need to take smaller chunks out of a larger data structure. Sometimes you will need to change the data that is stored in your structure, or use it as input to another function, or perhaps you need to plot it. 

The task of slicing a smaller chunk out of a larger data structure is called **indexing** in R (also  called **subsetting**), and is performed using the square bracket characters `[]`.

---

## Indexing 'Ingredients'

Indexing a data structure in R requires 4 ingredients. 

1. An R object that supports indexing 
    + e.g. A data frame, matrix, vector or list
2. An opening square bracket `[`
3. 1 (or more) vectors which indicate which values from the larger data structure should be pulled out. 
4. a closing square bracket `]`

Arrange the ingredients in an R expression like so:

`DataStructure[IndexVector]`

We will focus on learning what can go inside the square brackets for different types of R data structures

---

## Indexing Vectors

Broadly, there are 2 types of indexing vectors that are useful inside the square brackets. 

1. Numeric Vectors
    + e.g. `c(1, 4, 5, 10)`
2. Logical Vectors 
    + e.g. `c(TRUE,FALSE,FALSE,TRUE)`
    
We'll start with numeric indexing vectors to get a feel for the general procedure, and move up to logical indexing.

---

## Numeric Index Vectors 
We'll start by slicing smaller chunks out of a larger vector. Here, the numeric vectors inside the brackets tell R the *positions* of the elements we wish to extract.
```{r large_and_smaller}
alphabet <- c("a","b","c","d","e","f","g","h","i","j","k","l","m","n","o",
              "p","q","r","s","t","u","v","w","x","y","z") 
alphabet[c(1,26)] # Extract First and 26th element
alphabet[10:20] # Extract tenth through 20th
alphabet[seq(from=1,to=26,by=2)] # Extract every other element
```

---

## Common Errors with Numeric Indices
The most common mistake is including a value in your indexing vector which is greater than the length of the vector you are subsetting

```{r outofbounds, error=TRUE}
alphabet[100] # there are not 100 letters in the alphabet
```

We get an `NA` because the value is missing - there is no 100<sup>th</sup> letter of the alphabet. This is commonly referred to as an "index out of bounds" error.

Another common mistake is forgetting to concatenate the values you want to use for the indexing vector (i.e. forgetting the `c()` function).

.errormsg[
```{r no-c, error=TRUE}
alphabet[1,5,10]
```
]

---

## Indexing Tricks

Instead of creating a vector of values you *do* want to pick out, it may be easier to come up with a vector of ones you *don't* want. We can use negative number's to specify which vector elements we don't want. 

```{r negatives}
alphabet[c(-1,-26)] # Same as alphabet[2:24]
alphabet[-1:-10]
```

Indexing with positive vectors is usually preferred, as the intent of the code is more clear, but sometimes this is easier and more clear (e.g. when dropping the first or last value).

---

## Logical Indexing

When performing logical indexing, you supply a vector specifying whether to extract a specific element (with a `TRUE`) or to *not* extract a specific element (with a `FALSE`).

Let's revisit the example of selecting the first and last elements of the alphabet vector: We make a vector of logicals and stick it in the square brackets after your vector.

```{r bad_logical_indexing}
alphabet[c(1,26)]
alphabet[c(TRUE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE,
           FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE,
           FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE
           )
         ]
```

---

## Logical Indexing

But this specific example is not a good use case for logical vectors. Why?

1. Longer Code: length of the logical vector must match the length of the object its subsetting.
2. Duplicating work: If you already know the position of the elements you want, just put them into a numeric vector and you're done!

The utility of logical vectors occurs when you *don't* know the positions of the elements you are interested in, and must search for them!

---

## Logical Tests

"But wait" you say: "If we don't know where they are already, how are we going to find them?". This brings us to **logical testing** and **relational operators**

Relational operators are R expressions that test whether some value meets a condition or not.

- If the value meets your test's condition(s), the test returns `TRUE`
- If the value does *not* meet your test's condition(s), the test returns `FALSE`. 

You already know lots of relational operators. The equal to, greater than and less than expressions are all relational operators!

---

## Relational Operators Table

Comparison | Expression   | Example    | Result
---------- | :----------: | ---------- | 
Less Than | `<` | `5 < 10`<br>`5 < 1` | `r 5 < 10`<br>`r 5 < 1` 
Less Than or Equal To | `<=` | `5 <= 5`<br>`5 <= 1`| `r 5 <= 5`<br>`r 5 <= 1`
Greater Than | `>` | `10 > 5`<br>`5 > 5` | `r 10 > 5`<br>`r 5 > 5`
Greater Than or Equal To | `>=` | `10 >= 10`<br>`10 >= 12` | `r 10 >= 10`<br>`r 10 >= 12`
Equal To | `==` | `5 == 6`<br>`5 == 5` |  `r 5 == 6`<br>`r 5 == 5`
Not equal to | `!=` | `5 != 6`<br>`5 != 5` | `r 5 != 6`<br>`r 5 != 5`

---

## Relational Operators & Indexing
What makes relational operators useful is that they can be applied to all the elements of a data structure simultaneously. 
```{r lisa}
lisa <- c(34, 509, 63, 187, 998, 78, 3330)
print(lisa)
lisa <= 100 # Apply the "less than or equal to" test
```

As you can see, values that meet the criteria (<= 100) return as `TRUE`. When this logical vector is used to index the vector `lisa`, only the elements where the logical vector has value `TRUE` are returned.

```{r lisa_pt2}
lisa[lisa <= 100] # Index vector named lisa with the results of the test. 
```

---

## Relational Operators & Indexing
We index character vectors using the `==` and `!=` operators, but not the greater/less than operators. Quantity makes no sense for characters!
```{r charTesting}
months <- c("January", "February", "March", "April", "May", "June", "July",
           "August", "September", "October", "November", "December")
months == "June" # The sixth element is TRUE
months[months == "June"]
months[months != "July"]
```

---

## Other Useful Tests: `is.na()`
Unfortunately, we often have to deal with missing observations in real world data sets. R codes missing data as `NA` (or sometimes `NaN`). We can use the `is.na()` function to find any missing values in a vector. 
```{r isna}
missingno <- c(10,NA,1,4,2,NA,NA,99,NaN, NA)
is.na(missingno)
missingno[!is.na(missingno)] # Select things that are the opposite of missing
```

---

## Tips and Tricks
A useful function to know is `which()`. When used on a logical vector, it will return to your the position indices of the vector's `TRUE` element. It is useful when you want to know **where** in the vector your matches occur. 
```{r which}
lisa <= 100
which(lisa <= 100)
```

But don't do:
```{r bad_which}
lisa[which(lisa <= 100)]
```

Using `which` is an unneccesary extra step here that can slow performance.

---

## Testing Multiple Conditions 
Sometimes you need to select elements based on multiple conditions. For example, you might want to select only the rows of the `ncbirths` data where the mother was older than 35 *and* a smoker.

In R, we can select elements based on multiple conditions by combining individual logical tests together using *logical operators*. The logical operations we have at our disposal are: 

- AND
    - Each element must meet **all**  conditions to return `TRUE`
- OR
    - Each element must meet **at least one** conditions to return `TRUE`
- Negation
    - Reverse the current logical (e.g., `TRUE` becomes `FALSE`)

---

## Logical Operators 
Logic | Expression | Example | Result
----- | ---------- | ------- | ------
Elementwise AND | `&` | `c(1,3) > 0 & c(1,3) <=2` |`r c(1,3) > 0 & c(1,3) <=2` 
Elementwise OR | <code>&#124;</code> | <code>c(1,3) > 0 &#124; c(1,3) <=2</code> | `r c(1,3) > 0 | c(1,3) <=2` 
Negate | `!` | `!(c(1,3) > 0)` | `r !(c(1,3) > 0)`


The elementwise logical operators operators test all their arguments (i.e., they test all the pairs elements of the logical vectors supplied) and return a vector the same length as the input.

---

## Indexing with Multiple Conditions 
Let's say we wanted to select the elements in `lisa` which were less than 500 or greater than 1000.

```{r show_lisa_again}
lisa
```

To do this, we need to:

1. Test each element of `lisa` to see if it is less then 500
    - `lisa < 500`
2. Test each element of `lisa` to see if it is greater than 1000
    - `lisa > 1000`
3. Combine the results of both tests together into a single test
    - ???

---

## Indexing with Multiple Conditions
The goal is to have our test return a `TRUE` for each element that passes the "less than 500" test **OR** passes the "greater than 1000" test.

Expression | Element 1 | Element 2 | Element 3 | Element 4 | Element 5 | Element 6 | Element 7
------------- | --------- | --------- | --------- | --------- | --------- | --------- | ---------
`lisa` |     34    |   509     |    63     |    187    |    998    |     78    |    3330
`lisa < 500` | `TRUE` | `FALSE` | `TRUE` | `TRUE` | `FALSE` | `TRUE` | `FALSE`
`lisa > 1000` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `TRUE`
Pass Either Test? | <span style="color: Green">&#10003;</span> | 	 <span style="color: Red">&#10005; </span> | <span style="color: Green"> &#10003; </span> | <span style="color: Green"> &#10003; </span>| 	 <span style="color: Red"> &#10005; </span> | <span style="color: Green"> &#10003; </span>| <span style="color: Green"> &#10003; </span>

---

## Indexing with Multiple Conditions 
Our goal of the test is to have it return a `TRUE` for each element if it passed the "less than 500" test **OR** if it passed the "greater than 1000" test.

We can do that by combining the two tests with the `|` logical operator

Expression | Element 1 | Element 2 | Element 3 | Element 4 | Element 5 | Element 6 | Element 7
------------- | --------- | --------- | --------- | --------- | --------- | --------- | ---------
`lisa` |     34    |   509     |    63     |    187    |    998    |     78    |    3330
`lisa < 500` | `TRUE` | `FALSE` | `TRUE` | `TRUE` | `FALSE` | `TRUE` | `FALSE`
`lisa > 1000` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `FALSE` | `TRUE`
<code>lisa < 500 &#124; lisa > 1000</code> | `TRUE` | `FALSE` | `TRUE` | `TRUE` | `FALSE` | `TRUE` | `TRUE` 


```{r lisa_or}
lisa[lisa < 500 | lisa > 1000]
```

---

## Indexing with Multiple Conditions 
Now let's say we wanted to select the elements in `lisa` which were less than 500 and greater than or equal to 50.

To do this, we need to:

- Test each element of `lisa` to see if it is less then 500
     * `lisa < 500`
- Test each element of `lisa` to see if it is greater than or equal to 50
     * `lisa >= 50`
- Combine the results of both tests together into a single test
    * `&`

---

## Indexing with Multiple Conditions 
Our goal of the test is to have it return a `TRUE` for each element if it passed the "less than 500" test **AND** if it passed the "greater than 50" test.

Expression | Element 1 | Element 2 | Element 3 | Element 4 | Element 5 | Element 6 | Element 7
------------- | --------- | --------- | --------- | --------- | --------- | --------- | ---------
`lisa` |     34    |   509     |    63     |    187    |    998    |     78    |    3330
`lisa < 500` | `TRUE` | `FALSE` | `TRUE` | `TRUE` | `FALSE` | `TRUE` | `FALSE`
`lisa >= 50` | `FALSE` | `TRUE` | `TRUE` | `TRUE` | `TRUE` | `TRUE` | `TRUE`
`lisa < 500 & lisa >= 50` | `FALSE` | `FALSE` | `TRUE` | `TRUE` | `FALSE` | `TRUE` | `FALSE`

<br>
```{r lisa_and}
lisa[lisa < 500 & lisa >= 50]
```

---

## Indexing with Multiple Conditions 
```{r lisa_nothing}
lisa[lisa < 500 & lisa >= 998]
```
When you index any structure in R with a vector of all `FALSE`s, you get back `numeric(0)` which means "nothing to see here!"

So why does this test and subset return nothing?

Because no number can be less than 500 and greater than 998, so every test comes back `FALSE`, and all elements are ignored in the subset.

---

## Set-based indexing

Sometimes, you'll need to search for matches in a vector, but there are many possible matches that are acceptable.

For example, let's say we had a list of yes or no responses stored in a character vector, and we wanted to count up how many "yes" answers we got. Our first step would be to find each yes, and then count them from there.

But, people could respond "yes" in lots of ways. They might respond "yes", or "Yes", or "Y", or "y", or "YES", or maybe even "OK". If we want an accurate count, we'll need to accept *any* of these as a match, and also do the same thing for "no" responses.

```{r}
responses <- c("yes", "Yes", "no", "No", "y", "n", "NO", "Y", "YES")
```


What can we do?

---

## Set-based indexing

One thing we could do is chain together a bunch of equality tests with the OR operator.

```{r}
responses == "Y" | responses == "y" | responses == "YES" |
  responses == "yes" | responses == "Yes" | responses == "OK"
```

But there's a better way! R provides the `%in%` operator to look for matches in one set of values (like our vector of responses) anywhere in another set of values (like our set of possible "yes" answers).

```{r}
responses %in% c("yes", "Yes", "YES", "y", "Y", "OK")
```

We can read the `%in%` operator as answering the question "For each value on the left, is there a match anywhere on the right?"

---

## Indexing summary thus far...
- Indexing, or subsetting, is the process of slicing a smaller chunk out of a larger data structure
- Basic syntax template is `DataStructure[IndexVector]`
- the `IndexVector` can be a numeric or logical vector
- Logical testing uses relational operators like `<` and `==` and logical operators (e.g., `&` and `|`) to test which elements in our data structure meet specific requirements 


So far, we've explored this using vectors, which are one-dimensional data structure.

---

## Indexing Two-Dimensional Structures
Now we'll learn how to index data structures, such as data frames, that have two dimensions

As we've seen, data frames have both rows **and** columns, making them a 2D data structure

This means when we index them, we must specify which rows we would like to take out in our subset, as well as which columns

We'll practice this with the toy "sales" data frame we made earlier.
```{r print_earlier_df, ref.label="dataframeexample"}
```

---

## Indexing Data Frames
To select both rows and columns from a data frame, we just need two index vectors inside our square brackets, separated from each other by a comma.

The first vector applies to rows, the second applies to columns

The template is: `YourDataFrame[rowIndex, columnIndex]`

Just like when subsetting from vectors, the index vectors you supply can be either:

- numeric vectors specifying the row number(s) / column number(s)  you wish to select
- logical vectors specifying for each column and row whether we want to select it (`TRUE`)  or ignore it (`FALSE`)

---

## Data Frame Examples

```{r simple_df_indexing}
finances[1, 2:3] # Row 1, Column 2 and 3.
finances[1:2, 2:3] # Rows 1 and 2, Columns 2 and 3.  
```

---

## Data Frame Examples
You can select **all** of one dimension, (e.g., keep all rows or all columns) while indexing the other dimension. Just providing the separating comma as usual, but leave the space for the indexing vector blank  for the dimension you want to stay 100% intact.

```{r mat_examples3}
finances[1:2, ] # First two rows, all columns
finances[, 2:3] # All rows, second and third columns
```

---

## Name-based Indexing

Data Frames also support a powerful name-based set of indexing operations.

Indexing based on the name you've assigned to a row or column is almost always better, because that name is unlikely to change, while the row or column number is **very** likely to get changed

It's also much easier to remember the name of something than remember its position in a data frame.

---

## Name-based Indexing
If you want to select a single column by name, you can index the data frame using the `$` operator.

You start with the name of the data frame, follow it with the `$` operator, and finish with the *unquoted* name of the variable name.
```{r dollarsign}
finances$month
```

Note that the column is returned to you as a vector, *not* a data frame with one column.

```{r dropdim}
class(finances)
class(finances$month)
```
---

## Name-based Indexing

And remember, since this dollar-sign based indexing returns a vector, you can continue on indexing *that* vector itself!
```{r}
finances$month
finances$month[c(1,3)]
```

---

## Name-based Indexing
If you want to select *multiple* columns by name, you have to return to using the square bracket technique. In this case, you provide the list of column names you want to select as a character vector

```{r}
finances[, c("sales", "month")]
```

You're free to combine position-based indexing on the rows, and name-based indexing on the columns
```{r}
finances[1:2, c("sales", "month")]
```

---

## Logical Indexing & 2D structures
One of the most common subsetting tasks with a data frame is to select values in one column where the values in *another* column meet a certain criteria.

For example, perhaps we want to find out which month we had our best sales figures in. So, we need to find the row(s) in our data set where the `sales` variable is equal to the *maximum* of the `sales` variable.

We can break this job down into a three-step process:

1. Find the maximum value in the `sales` variable
2. Test each value in the `sales` variable to see if it equals this maximum number
3. Select just the rows of the data set that pass this test

---

## Logical Indexing & 2D structures

Step 1 can be solved by using the `max()` function on the `sales` variable.
```{r}
max_sales <- max(finances$sales)
max_sales
```

Step 2 can be solved by comparing each value in the `sales` variable to the `max_sales` value using the `==` comparison operator.
```{r}
finances$sales == max_sales
```

Step 3 can be solved by indexing the rows of the `finances` data frame using the logical vector that is returned from step 2.
```{r}
finances[finances$sales == max_sales, ]
```

---

## Practice
Which months had costs over $200?

Can we make sure *only* the months are printed out, and not any sales or costs values?

Which months did we make a profit? In other words, in which months were sales greater than profits?

---

## Errors when indexing by name
If you try to select a column of a data frame using the `$` operator, but there is no column with that name, `R` will return `NULL`

```{r NULL}
finances$profits
```

But, if you use the `[row, column]` style of indexing and ask for a column that doesn't exist, you get a proper error. 

```{r col_error, error=TRUE}
finances[,'profits']
```

---

## Editing data frames
You can use indexing operation on the left hand side of an assignment operation to remove or replace values in your data structure. The basic recipe looks like:

`DataFrame[LogicalCriteria] <- NewValues`

A note of caution: this is an irreversible operation. It would behoove you to make a backup copy of your data structure before altering it, like so:

`backup_object <- DataObject`

`DataObject[LogicalCriteria] <- NewValues`

---

## Editing data frames
Let's say that we made a mistake when entering our sales data, and we actually made $500 in January! Let's update our data frame to reflect the good news! To do this, we need to:

1. Search for the "Jan" row in the data frame
2. Use the results of this search to select the "Jan" row of the data frame, as well as the "Sales" column
3. Overwrite the selected cell of the data frame with the number 500

--

```{r}
finances[finances$month == "Jan", "sales"] <- 500
finances
```
            ↑
           Tada!
---

## Removing rows

If you want to remove rows, it's generally more effective to approach the problem as choosing which rows to keep, rather than which rows to delete.

For example, if we wanted to remove the month of September from our financial data, we wouldn't "delete" that row - rather, we would make a new data frame the keeps every row *except* for the September

```{r remove_sep}
finances[finances$month != "Sep",]
```

---

## Removing columns
You can remove a *single* column by indexing it with the `$` operator, and assigning it the special value `NULL`.

```{r}
finances$month <- NULL
finances
```

If you want to remove *multiple* columns, it's best to think about it the same way as rows - make a new data frame that keeps all the columns you want, and none of the ones you want gone, instead of deleting columns one by one.

