
---
title: 'Problem Set 04: Data Manipulation with dplyr'
author: "YOUR NAME HERE =)"
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float: yes
    toc_level: 2
    df_print: kable
    self_contained: true
editor_options: 
  chunk_output_type: console
---

<!-- PLEASE DONT TOUCH ANY OF THE STYLING CODE BELOW -->
<style type="text/css">
body {
  counter-reset: exercises toc; /* initialize counters */
  font-size: 18px;
}
.col-md-3 { width: 12%; }

#TOC { width: 65%; max-width: 130px; }

.col-md-9 { width: 88%; }

p { margin: 5px 0;}

img {margin: 5px 0;}

ol li, ul li { list-style-position: inside; }

div[id^='exercise'] { background:rgba(49, 126, 172, 0.1); }

div[id^='exercise'] > h2 {
  font-weight: bold;
  color: #fff;
  background:#317EAC;
  padding: 5px 10px;
}

div[id^='exercise'] > h2::after {
  content: " " counter(exercises);
  counter-increment: exercises;
}

.tocify-header > li:after {
    content: " " counter(toc); /* Use the counter as content */
    counter-increment: toc; /* Increment the counter by 1 */
}

.level3 {
  background: #fff;
  padding-top: 15px;
  margin-bottom: 0;
}

.level3 > h3 { margin: 0;}

.attention { 
  color: #000;
  border: 2px solid #ff2e2e;
  padding: 10px 10px;
  background-color: rgb(255, 110, 110);
  border-radius: 5px;
}
</style>

<!-- Exercises begin below -->

In this problem set, we will practice some of the key data manipulation tasks using the tools from the `dplyr` package.

### Libraries
```{r warning = F, message = F}
library(ggplot2)
library(dplyr)
```

### The data
You will be working with the `txhousing` data set, which is included in the `ggplot2` package and will be loaded when you use `library(ggplot2)`. Take a peek at the dataset below, taking note of the data types of each variable:

```{r R.options=list(tibble.width=85)}
glimpse(txhousing)
```

This data set has **monthly** observations of the housing market in 46 regions of Texas from the years 2000 through 2015. The `city`, `year`, `month` and `date` variables identify the city, month, year, and exact date of the observation, and the measured variables are for each observation are:

- `sales`: Number of sales
- `volume`: Total value of sales
- `median`: Median sale price
- `listings`: Total number of homes listed for sale
- `inventory`: A "Months inventory", a.k.a, the amount of time it would take to sell all current listings at current pace of sales.

<div class="attention">
For each problem below, always print out the final data frame or vector that gives you the answer to the question.
</div>

## Exercise
Remove the `date` variable from the data frame, rename the `median` variable to `median_price`. Be sure to overwrite the original data frame with your results.

### Solution


## Exercise
Remove any rows of the `txhousing` data set with missing values in the `sales` variable, and overwrite the original data frame. Then use the `anyNA()` function on the `sales` variable, e.g., `anyNA(txhousing$sales)`. If you were successful, this should print out `FALSE`.

### Solution


## Exercise
Make a data set called `dallas` that includes data only from the city of Dallas in the years 2000 through 2010. Test to see if your results are correct by using the `distinct` function to show the unique city and year combinations in your data frame.

### Solution


## Exercise
What are the median sales volume and the total sales volume for homes in Dallas, across all months and years in the `dallas` data frame? Your solution should present both answers in the same data frame (no need to save the output here). Remember, you can compute the median value with the `median` function.

### Solution


## Exercise
Compute the total sales volume for each year in the `dallas` data frame. Save this output as a new data frame called `dallas_annual_volume`.

*Hint: The correct solution will create a data frame with 11 rows; one row for 2000, one row for 2001, one row for 2002, etc.)*

### Solution


## Exercise:
Which year in the had the lowest sales volume in the  `dallas_annual_volume` data frame? Your solution should print out the single row of the data frame that corresponds to this year. No need to save the results.

### Solution


## Exercise
Using `ggplot`, visualize the sales volume per year from the `dallas_annual_volume` data frame. Describe the pattern of sales volumes over time.

**Hints: Any variable representing time usually goes on the x axis of a figure. You can use the geom_point() function to draw individual points on a plot, and the `geom_line() function to connect the points**

### Solution 


## Exercise
Add a column to the `txhousing` data set called `houses_left` that calculates the difference between the number of homes listed for sale and the number of homes actually sold at the end of the month. Be sure to save the `txhousing` data frame with this new column. 

### Solution

## Exercise
Which city had the highest sales volume over the entire period recorded in the `txhousing` dataset? Which city had the most sales? Are they the same city?


### Solution


## Exercise
Generate a single table that shows the total number of houses sold in Austin and Dallas over the period of years 2000 and 2001.

### Solution

